```{r global_options, include=FALSE}
library(knitr)
opts_chunk$set(fig.align="center", warning=FALSE, message=FALSE, idy.opts=list(width.cutoff=120))
```

# Introduction to alignfigR

<br>

The package **alignfigR**, built around ggplot2, creates multiple sequence alignment figures in R. Source code is freely available on github: [https://github.com/sjspielman/alignfigR](https://github.com/sjspielman/alignfigR). Importantly, the alignment plot that alignfigR returns to you is also a ggplot object, which you can subsequently label, mark up, and save to your heart's content.

<br>

### Installation

alignfigR can be installed using the R library devtools (which you'll need to install first if you don't have already):
```{r message=FALSE}
library(devtools)
devtools::install_github("sjspielman/alignfigR")
library(alignfigR)
```

<br>

### General usage

First, use the function `read_alignment()` to load sequence data into R. Currently, only FASTA format is supported.
```{r message=FALSE}
my_data <- read_alignment("../data/example.fasta")
```

<br>

Then, simply create an alignment figure with the function `plot_alignment()`. This function takes two ordered arguments: 

1. *alignment*, which corresponds to your sequence data set
2. *palette*, which gives the color-mapping scheme scheme for your sequence data. Several options exist for this argument, as follows.

```r
# Default DNA colors
p <- plot_alignment(my_data, "dna")

# Default RNA colors
p <- plot_alignment(my_data, "dna")

# Default protein colors
p <- plot_alignment(my_data, "protein")

# Random colors are the default. Either provide the argument "random", or simply provide nothing.
p <- plot_alignment(my_data)
p <- plot_alignment(my_data, "random")

```

<br>

You can also specify your own color scheme using a named-array, as follows. Note that missing characters (such as gaps) can also be colored. This option is particularly useful for dealing with noncanonical data (e.g. binary or character data). However, word of caution! If any alignment chartacters are not assigned a color, such characters will be left as whitespace in the resulting plot. 
```r
my_favorite_colors <- c("A" = "pink", "C" = "magenta", "G" = "seagreen", "T" = "yellow",
                        "-" = "black")
p <- plot_alignment(my_data, my_favorite_colors)
```

<br>

Here, we will use the default color scheme for a DNA alignment.

```{r basic_dna, message=FALSE,  fig.height=2.5, fig.width=5.5}
p <- plot_alignment(my_data, "dna")
p
```

<br>

As mentioned, you can manipulate this figure in any way you want, using ggplot2, from here on out. For instance, maybe a title!
```{r titled_basic_dna, tidy=TRUE, tidy.opts=list(width.cutoff=120), echo=TRUE, fig.height=2.5, fig.width=5.5}
library(ggplot2)
p + ggtitle("My fancy-schmancy alignment figure!") 
```


<br><br>

### Plotting alignment subsets


By default, `plot_alignment()` will create a figure for your entire alignment. However, it is also possible to plot only a subset of your alignment, selecting particular taxa and/or columns. 

<br>

To restrict the plot to certain taxa, specify the taxa you'd like to keep with the argument "taxa".
```{r dna_restrict_taxa, tidy=TRUE, tidy.opts=list(width.cutoff=120), fig.height=2.5, fig.width=5.5}
plot_alignment(my_data, "dna", taxa = c("Cow", "Carp"))

```

<br>

You can alternatively exclude specified taxa from the plot by adding the argument `exclude_taxa = T`.
```{r dna_exclude_taxa, tidy=TRUE, tidy.opts=list(width.cutoff=120), fig.height=2.5, fig.width=5.5}
plot_alignment(my_data, "dna", taxa = c("Cow", "Carp"), exclude_taxa = T)
```

<br>

Columns can be similarly specified with the argument "columns".
```{r dna_restrict_cols, tidy=TRUE, tidy.opts=list(width.cutoff=120), fig.height=2.5, fig.width=5.5}
plot_alignment(my_data, "dna", columns = c(1:25))
```

<br>

And again, you can instead exclude specific columns by adding the argument `exclude_columns = T`.
```{r dna_exclude_cols, tidy=TRUE, tidy.opts=list(width.cutoff=120), fig.height=2.5, fig.width=5.5}
plot_alignment(my_data, "dna", columns = c(1:200), exclude_columns = T)
```

<br>

And of course, we can also combine these options do get any alignment subset we want (with exciting colors, too)!
```{r dna_colored_restrict_cols_taxa, tidy=TRUE, tidy.opts=list(width.cutoff=100), fig.height=2.5, fig.width=5.5}
exciting_colors <- c("A" = "turquoise", "C" = "maroon", "G" = "mediumpurple1", "T" = "royalblue4", "-" = "cornsilk1")
plot_alignment(my_data, exciting_colors, columns = c(1:200, 350:450), exclude_columns = T, taxa = c("Cow", "Carp", "Chicken", "Human") )
```







